Ecosystem size and complexity dictate riverine biodiversity

### 2021-03-01 manuscript_science.Rmd
edits: re-format to Science

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro, Greensboro, NC 27412, USA]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of
  Minnesota, St. Paul, MN 55108, USA]
- Taku Kadoya^[National Institute for Environmental Studies, Tsukuba, Ibaraki 305-8506, Japan]
- Yusuke Miyazaki^[Shiraume Gakuen College, Kodaira, Tokyo 187-8570, Japan]
output: word_document
abstract: The species-area relationship, a tendency in which species richness increases with increasing area, is a ubiquitous pattern in ecology[@lomolino2000] and is the foundation of many conservation strategies[@myers2000; @desmet2004; @neigel2003]. However, many ecosystems possess a complex spatial structure that cannot be represented simply by area[@rodríguez-iturbe2001; @turner2015], and the role of such scale-invariant complexity in scaling biodiversity relationships is largely unknown. Here, we use theory and empirical analysis of extensive fish community data to show that ecosystem size and complexity jointly regulate biodiversity in rivers. Rivers are a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns[@rodríguez-iturbe2001]. Ecosystem complexity, which we define as the probability of branching per unit river distance[@terui2018; @yeakel2014], should increase watershed-scale species richness by increasing habitat heterogeneity. Meanwhile, ecosystem size should dictate metacommunity size and total habitat diversity, two factors that regulate watershed-scale species richness[@hubbell2001; @kallimanis2008]. Metacommunity simulations in branching networks predicted biodiversity patterns consistent with our hypothesis. Larger and more branched ecosystems supported greater species richness at the watershed scale, and the importance of local environmental filtering of species was increasingly apparent when simulated species dispersal was limited. Cross-watershed comparisons across two geographic regions (Hokkaido Island in Japan and the Midwestern US) provided empirical support for the theoretical predictions. In both regions, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial difference in species composition. Recognizing the dual control of biodiversity should broaden viable options for spatial planning of protected areas or restoration, thereby helping conserve biodiversity from societal demands that threaten it.

csl: nature.csl
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
---

@@

@@

Ecologists have long sought to understand the general drivers of biodiversity. One of the most robust empirical generalizations in ecology is the positive relationship between species richness and area, i.e., the species-area relationship (the SAR)[@lomolino2000]. In 1921, Arrhenius[@arrhenius1921] formulated the SAR as a power-law $S = cA^z$, an equation currently known as the Arrhenius species-area relationship ($S$ is the number of species observed in a given geographic area $A$, $c$ the constant, and $z$ the scaling exponent). Since then, the spatial scaling of species richness has been observed in many taxonomic groups[@triantis2012]. The SAR is ubiquitous because multiple mechanisms produce an apparently similar pattern. Larger ecosystems typically support more diverse metacommunities due to increased habitat diversity[@kallimanis2008], larger metacommunity size[@hubbell2001], and/or enhanced colonization dynamics[@losos2009]. Importantly, the SAR provides the foundation for global conservation efforts[@myers2000; @desmet2004; @neigel2003]. For example, conservation ecologists have used SAR estimates to design marine and terrestrial protected areas[@desmet2004; @neigel2003], which currently encompass more than 30 million km^2^ globally[@deguignet2014].

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity[@turner2015; @rodríguez-iturbe2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few[@rodríguez-iturbe2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike[@rodríguez-iturbe2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms[@grant2007]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure, and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems[@rinaldo2018]. For example, these studies have highlighted key roles of branching structure in driving local biodiversity patterns, such as increased species richness at merging points of branches[@carrara2012]. However, most empirical research has explored the consequences of branching complexity for local community structure[@ers2019; @carrara2012] or has relied solely on theoretical arguments with limited replications of branching architecture[@holt2018]. At present, we lack a comprehensive evaluation of how branching complexity scales biodiversity at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where accelerated species loss threatens the delivery of ecosystem services[@dudgeon2006].

Here, we hypothesize that ecosystem size and complexity dictate biodiversity patterns in rivers - a prime example of complex branching ecosystems. Individual streams and rivers flow through different landscapes with distinct geological and climatic backgrounds, serving as a spatial unit of unique in-stream environments[@peterson2013; @moore2015; @terui2018]. The recurrent merging of diverse tributaries ultimately forms a fluvial network with fractal branching patterns[@rodríguez-iturbe2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance[@terui2018; @yeakel2014], may represent a macro-scale control of the ecosystem's habitat heterogeneity (habitat diversity per unit area)[@grant2007; @moore2015; @terui2018]. Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity), two factors that regulate biodiversity at the metacommunity level[@hubbell2001; @iwasaki2012]. We predict that ecosystem size and branching complexity increase watershed-scale species richness ($\gamma$ diversity) by enhancing local species richness ($\alpha$ diversity) and/or spatial difference of species composition ($\beta$ diversity) under different ecological scenarios. The present study combines theory and empirical analysis of extensive community data from two different regions of the globe to provide crucial insights into how ecological communities are structured in complex branching networks.

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We depicted branching ecosystems as a spatial network of connected habitat patches which local communities inhabit[@terui2018; @yeakel2014] (**Figure 1A and B**). Two factors determine mean environmental conditions at each habitat patch: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see **Methods**). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We then simulated metacommunity dynamics in the theoretical branching networks using a general metacommunity framework[@thompson2020]. In this model, 50 species with different abiotic niches (optimum and width) disperse along a network and compete for resources with varied strengths in temporally dynamic environments (see **Methods** for model details).

Given the results of extensive sensitivity analysis (**Supplementary Information**), we considered `r n_scenario` simulation scenarios comprising a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation (**Figure 1A**). This landscape scenario should reproduce natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries but are highly correlated within a tributary[@moore2015; @peterson2013]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to limited influences of branching on habitat heterogeneity (**Figure 1B**). This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities[@poff2007; @terui2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel $\theta$), dispersal probability, and the maximum value of interspecific competition strength $\alpha_{max}$ (see Methods). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (**Supplementary Information**). Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 400 time steps of initialization and burn-in periods) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

However, dispersal processes affected mechanisms that underlie the positive effects of ecosystem size and complexity on $\gamma$ diversity (compare left and right columns in **Figures 2 and 3**). We observed a greater contribution of $\beta$ diversity (defined as $\frac{\gamma}{\alpha}$) to increased $\gamma$ diversity when dispersal limitation existed (i.e., species travel short distances). This pattern reflected significant spatial variation in species composition over the branching network and was likely driven by the local association of species' niche and abiotic environments (i.e., species sorting)[@thompson2020]. In contrast, when the dispersal limitation was relaxed (species travel long distances), a clear increase in $\alpha$ diversity underpinned the positive relationships between $\gamma$ diversity and ecosystem properties. The results agree with previous predictions that increased dispersal homogenizes community composition while enhancing local diversity through increased immigrants from suitable habitat patches (i.e., mass effects)[@thompson2020]. These patterns were consistent across different levels of dispersal probabilities (**Figures S5 and S12**). The strength of competition decreased the maximum of $\alpha$ diversity but did not change the scaling relationships with ecosystem properties (**Figures 2 and 3**). In summary, our theory highlights how apparently similar patterns in $\gamma$ diversity emerge through different ecological pathways.

Influences of ecosystem size and complexity differed significantly in their dependence on landscape scenarios. Ecosystem size had positive effects on $\gamma$ diversity regardless of landscape scenarios, although the slopes were steeper with greater environmental variation (higher $\sigma_h$ and/or $\sigma_l$) (**Figures 2 and S5-11**). This result is attributable to the fact that larger ecosystems can hold more individuals in a metacommunity[@hubbell2001]. In contrast, we observed limited or no influences of branching complexity when local environmental noise was equal to or exceeded environmental variation at headwaters ($\sigma_l \ge \sigma_h$; **Figures S13-18**). Under this scenario, branching has a minor influence on the ecosystem's habitat heterogeneity because patch-level environmental variation is equivalent to or greater than environmental differences between tributaries (**Figure 1B**). Therefore, this landscape scenario eliminates the positive effect of branching complexity on $\gamma$ diversity. This theoretical prediction may not apply to pristine or semi-natural river networks where individual streams show distinct and spatially-correlated environmental conditions, including water temperature, water chemistry, and flow/sediment regimes[@peterson2013; @moore2015]. Instead, it may be most relevant to severely altered landscapes where localized human disturbance disrupts the environmental distinctiveness of branches through, for example, flow regulations by dams[@poff2007]. Hence, our theory has important implications for riverine biodiversity conservation by pointing to the crucial role of habitat diversity produced by branching structure.

The proposed theory provided important insights into how ecological communities are structured in branching networks; however, empirically testing the predictions is extremely difficult because it requires metacommunity-level replications. To confront this logistical challenge, we compiled fish community data across two geographic regions: Hokkaido Island in Japan and the Midwestern US (Midwest). These regions are located in comparable latitude ranges (**Figure 1D**) but support distinct fish communities (**Tables S5 and S6**). Therefore, this data set provides an excellent opportunity to examine the generality of our theoretical predictions. After careful data selection (see **Methods** and **Supplementary Information**), we estimated $\alpha$, $\beta$, and $\gamma$ diversity (asymptotic species richness; **Methods**) at `r nrow(wsd_subset[[1]]) + nrow(wsd_subset[[2]])` watersheds (`r nrow(wsd_subset[[1]])` in Hokkaido and `r nrow(wsd_subset[[2]])` in the Midwest), each of which comprised $\ge$ 10 sites of presence-absence fish community data (a total of `r nrow(point_subset[[1]]) + nrow(point_subset[[2]])` sites). We considered watersheds as independent replicates of metacommunities if the river network flows into one of the following: the ocean, a large lake/reservoir ($\ge$ 10 km² in the areal area), or a large river that may represent lentic habitats ($\ge$ 5000 km² in the watershed area). These watersheds are small enough (`r round(min(c(wsd_subset[[1]]$area, wsd_subset[[2]]$area)))` to `r round(max(c(wsd_subset[[1]]$area, wsd_subset[[2]]$area)))` km^2^) to assume that fishes can disperse therein at a multi-generation time scale[@comte2018], while posing challenges to traverse across watersheds (the ocean or lentic habitats). We combined this data set with geospatial information, including watershed characteristics (watershed area and branching probability), climates (annual mean temperature and annual cumulative precipitation), and land use patterns (the fraction of agricultural land use and dam density). Using this data set, we assessed whether observed effects of the macro-scale factors are consistent across the two geographic regions by developing global and region-specific models. In the global model, we assumed that effects of ecosystem size (watershed area) and complexity (branching probability) are constant across the two regions (i.e., fixed slopes). Meanwhile, the region-specific model assumes region-specific slopes of ecosystem size and complexity (**Methods**). We compared the support of these competing models using the Bayes factor, a measure of the strength of evidence in favor of one model over the alternative. In our definition (**Methods**), the Bayes factor of $> 1$ indicates the support for the global model over the region-specific model.

Despite the substantial difference in fish fauna between the study regions, we found strong support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`): this result suggests the consistent effects of ecosystem size and complexity on $\gamma$ diversity across the two geographic regions. Furthermore, the observed patterns were closely aligned with our theoretical predictions (**Table 1 and Figure 4**). The estimated $\gamma$ diversity increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) across watersheds. These effects remained significant even after controlling for the potential influences of other environmental factors (precipitation, temperature, agricultural land use, and dam density) (**Table 1**). Our statistical analysis provides the first empirical evidence that ecosystem complexity regulates riverine biodiversity at the metacommunity level while conferring additional support for the ecosystem size effect on fish communities[@iwasaki2012].

Similarly, we found weak to moderate supports for the global models of $\alpha$ and $\beta$ diversity (Bayes factor = `r round(fit$alpha$bayes_factor, 1)` and `r round(fit$beta$bayes_factor, 1)`, respectively). In both regions, $\beta$ diversity responded significantly to ecosystem size and complexity while $\alpha$ diversity showing a weaker or a vague response to these ecosystem properties (**Table 1 and Figure 4**). In our simulations, this pattern has emerged under the scenarios with dispersal limitation, which elegantly matches the previous observations of stream fish movement. Field studies (e.g., mark-recapture) recurrently revealed the restricted movement of stream fish, typically limited to several tens to hundreds of meters at an annual scale[@radinger2014]. The reciprocal agreement of theoretical and empirical patterns provides indirect but convincing evidence that dispersal limitation plays a key role in driving the associations between $\gamma$ diversity and ecosystem properties in rivers.

@@

The emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation[@poff2007], habitat fragmentation[@grill2019], and stream burial[@elmore2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research[@koning2020]. While large reserves or restoration sites are undoubtedly important, our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin[@koning2020].

While the prevailing evidence supports the importance of ecosystem size in scaling species diversity[@hubbell2001; @triantis2012; @iwasaki2012], ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial[@turner2015] and aquatic ecosystems[@rodríguez-iturbe2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings of the dual control of biodiversity should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. The dual dimensional approach - size and complexity - broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

# Methods

## Theory

### Network generation

We depicted branching ecosystems as a bifurcating network of connected habitat patches[@terui2018; @yeakel2014] (**Figure 1**). Habitat patches, or local communities, can be either non-branching or branching river sections with a unit length $l$, which defines the scale of local species interactions. Two parameters determine the geometric properties of simulated branching networks: the number of habitat patches $N_p$ (ecosystem size) and branching probability $P_b$ (ecosystem complexity). Each habitat patch is assigned to be a branching patch (including upstream terminals) with probability $P_b$ or a non-branching patch with probability $1-P_b$. In this framework, an individual branch is a consecutive series of non-branching patches terminated at a branching patch; therefore, the number of habitat patches in branch $q$, $n_{p,q}$, is a realization of a random variable drawn from a geometric distribution $n_{p,q} \sim Ge(P_{b})$[@terui2018] but conditional on $\sum_{q=1}^{N_b}n_{p,q} = N_p$ ($N_b$ is the number of branches in a network). The number of branches $N_b$ is identical to the number of branching habitat patches in a network, or the number of 'success' out of $N_p$ trials. This leads to an assumption that $N_b$ is a realization of a random variable drawn from a binomial distribution as $N_b \sim Binomial(N_p, P_b)$. This framework has two merits. First, it reflects observed patterns of branch length distribution, which is known to follow an exponential distribution (a continuous version of a geometric distribution)[@rodríguez-iturbe2001]. Second, it preserves the fractal nature of branching patterns[@peckham1999]. An alternative approach is the use of optimal channel networks[@carraro2020]. However, we employed our framework because branching probability $P_b$ is quantifiable in natural river networks, enabling us to compare theoretical predictions with empirical observations.

In our network generation, we first drew $N_b$ from the binomial distribution but selected an odd number to constitute a bifurcating branching network. We then determined $n_{p,q}$ as random draws from the geometric distribution with the constraint of $\sum_{q=1}^{N_b}n_{p,q} = N_p$. The generated branches were organized randomly into a bifurcating branching network.

>>>>

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
output:
  pdf_document:
    latex_engine: xelatex
csl: science.csl
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
---

@@

**One sentence summary:** Using theory and empirical analysis of extensive community data, this research provides the first robust evidence that ecosystem size and complexity jointly regulate biodiversity in rivers at large spatial scales.

**Authors:** Akira Terui^1^, Seoghyun Kim^1^, Christine L. Dolph^2^, Taku Kadoya^3^, Yusuke Miyazaki^4^

**Affiliations:** ^1^ Department of Biology, University of North Carolina at Greensboro, Greensboro, NC 27412, USA ^2^ Department of Ecology, Evolution and Behavior, University of Minnesota, St. Paul, MN 55108, USA ^3^ National Institute for Environmental Studies, Tsukuba, Ibaraki 305-8506, Japan ^4^ Shiraume Gakuen College, Kodaira, Tokyo 187-8570, Japan

**Abstract:** Larger ecosystems generally support more species; this ubiquitous ecological pattern is the foundation of many conservation strategies. However, many ecosystems possess a complex spatial structure that cannot be represented by area, and the role of such ecosystem complexity in scaling biodiversity is largely unknown. Here, we use theory and empirical analysis of extensive fish community data from two geographic regions (Japan and US) to show that ecosystem size and complexity dictate riverine biodiversity. We found that larger and more branched 'complex' river networks harbored greater species richness due to increased space and environmental heterogeneity. The complexity effect was comparative to the size effect, and this pattern emerged regardless of ecological contexts. Recognizing the dual control of biodiversity should broaden viable options for successful conservation.

@@

Ecologists have long sought to understand the general drivers of biodiversity. One of the most robust empirical generalizations in ecology is the positive relationship between species richness and area, i.e., the species-area relationship (the SAR) [@lomolino2000]. In 1921, Arrhenius [@arrhenius1921] formulated the SAR as a power-law $S = cA^z$, an equation currently known as the Arrhenius species-area relationship ($S$ is the number of species observed in a given geographic area $A$, $c$ the constant, and $z$ the scaling exponent). Since then, the spatial scaling of species richness has been observed in many taxonomic groups [@triantis2012]. The SAR is ubiquitous because multiple mechanisms produce an apparently similar pattern. Larger ecosystems typically support more diverse metacommunities due to increased habitat diversity [@kallimanis2008], larger metacommunity size [@hubbell2001], and/or enhanced colonization dynamics [@losos2009]. Importantly, the SAR provides the foundation for global conservation efforts [@myers2000; @desmet2004; @neigel2003]. For example, conservation ecologists have used SAR estimates to design marine and terrestrial protected areas [@desmet2004; @neigel2003], which currently encompass more than 30 million km^2^ globally [@deguignet2014].

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity [@turner2015; @rodríguez-iturbe2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few [@rodríguez-iturbe2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike [@rodríguez-iturbe2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms [@grant2007]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure, and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems [@rinaldo2018]. For example, these studies have highlighted key roles of branching structure in driving local biodiversity patterns, such as increased species richness at merging points of branches [@carrara2012]. However, most empirical research has explored the consequences of branching complexity for local community structure [@ers2019; @carrara2012] or has relied solely on theoretical arguments with limited replications of branching architecture [@holt2018]. At present, we lack a comprehensive evaluation of how branching complexity scales biodiversity at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where accelerated species loss threatens the delivery of ecosystem services [@dudgeon2006].

Here, we hypothesize that ecosystem size and complexity dictate biodiversity patterns in rivers - a prime example of complex branching ecosystems. Individual streams and rivers flow through different landscapes with distinct geological and climatic backgrounds, serving as a spatial unit of unique in-stream environments [@peterson2013; @moore2015; @terui2018]. The recurrent merging of diverse tributaries ultimately forms a fluvial network with fractal branching patterns [@rodríguez-iturbe2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@terui2018; @yeakel2014], may represent a macro-scale control of the ecosystem's habitat heterogeneity (habitat diversity per unit area) [@grant2007; @moore2015; @terui2018]. Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity), two factors that regulate biodiversity at the metacommunity level [@hubbell2001; @iwasaki2012]. We predict that ecosystem size and branching complexity increase watershed-scale species richness ($\gamma$ diversity) by enhancing local species richness ($\alpha$ diversity) and/or spatial difference of species composition ($\beta$ diversity) under different ecological scenarios. The present study combines theory and empirical analysis of extensive community data from two different regions of the globe to provide crucial insights into how ecological communities are structured in complex branching networks.

# Synthesizing ecosystem size and complexity influences

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We depicted branching ecosystems as a spatial network of connected habitat patches which local communities inhabit [@terui2018; @yeakel2014] (**Figure 1A and B**). Two factors determine mean environmental conditions at each habitat patch: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see **Materials and Methods**). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We then simulated metacommunity dynamics in the theoretical branching networks using a general metacommunity framework [@thompson2020]. In this model, 50 species with different abiotic niches (optimum and width) disperse along a network and compete for resources with varied strengths in temporally dynamic environments (see **Materials and Methods** for model details).

Given the results of extensive sensitivity analysis (**Supplementary Text**), we considered `r n_scenario` simulation scenarios comprising a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation (**Figure 1A**). This landscape scenario should reproduce natural patterns of habitat heterogeneity, in which environmental conditions differ greatly among tributaries but are highly correlated within a tributary [@moore2015; @peterson2013]. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to limited influences of branching on habitat heterogeneity (**Figure 1B**). This scenario may reflect human-modified landscapes where the physical or biological distinctiveness of tributaries has been compromised due to human activities [@poff2007; @terui2018]. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity within a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel $\theta$), dispersal probability, and the maximum value of interspecific competition strength $\alpha_{max}$ (see **Materials and Methods**). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (**Supplementary Text**). Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 400 time steps of initialization and burn-in periods) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

However, dispersal processes affected mechanisms that underlie the positive effects of ecosystem size and complexity on $\gamma$ diversity (compare left and right columns in **Figures 2 and 3**). We observed a greater contribution of $\beta$ diversity (defined as $\frac{\gamma}{\alpha}$) to increased $\gamma$ diversity when dispersal limitation existed (i.e., species travel short distances). This pattern reflected significant spatial variation in species composition over the branching network and was likely driven by the local association of species' niche and abiotic environments (i.e., species sorting) [@thompson2020]. In contrast, when the dispersal limitation was relaxed (species travel long distances), a clear increase in $\alpha$ diversity underpinned the positive relationships between $\gamma$ diversity and ecosystem properties. The results agree with previous predictions that increased dispersal homogenizes community composition while enhancing local diversity through increased immigrants from suitable habitat patches (i.e., mass effects) [@thompson2020]. These patterns were consistent across different levels of dispersal probabilities (**Figures S5 and S12**). The strength of competition decreased the maximum of $\alpha$ diversity but did not change the scaling relationships with ecosystem properties (**Figures 2 and 3**). In summary, our theory highlights how apparently similar patterns in $\gamma$ diversity emerge through different ecological pathways.

Influences of ecosystem size and complexity differed significantly in their dependence on landscape scenarios. Ecosystem size had positive effects on $\gamma$ diversity regardless of landscape scenarios, although the slopes were steeper with greater environmental variation (higher $\sigma_h$ and/or $\sigma_l$) (**Figures 2 and S5-11**). This result is attributable to the fact that larger ecosystems can hold more individuals in a metacommunity [@hubbell2001]. In contrast, we observed limited or no influences of branching complexity when local environmental noise was equal to or exceeded environmental variation at headwaters ($\sigma_l \ge \sigma_h$; **Figures S13-18**). Under this scenario, branching has a minor influence on the ecosystem's habitat heterogeneity because patch-level environmental variation is equivalent to or greater than environmental differences between tributaries (**Figure 1B**). Therefore, this landscape scenario eliminates the positive effect of branching complexity on $\gamma$ diversity. This theoretical prediction may not apply to pristine or semi-natural river networks where individual streams show distinct and spatially-correlated environmental conditions, including water temperature, water chemistry, and flow/sediment regimes [@peterson2013; @moore2015]. Instead, it may be most relevant to severely altered landscapes where localized human disturbance disrupts the environmental distinctiveness of branches through, for example, flow regulations by dams [@poff2007]. Hence, our theory has important implications for riverine biodiversity conservation by pointing to the crucial role of habitat diversity produced by branching structure.

# Empirical evidence from distinct geographic regions

The proposed theory provided important insights into how ecological communities are structured in branching networks; however, empirically testing the predictions is extremely difficult because it requires metacommunity-level replications. To confront this logistical challenge, we compiled fish community data across two geographic regions: Hokkaido Island in Japan and the Midwestern US (Midwest). These regions are located in comparable latitude ranges (**Figure 1D**) but support distinct fish communities (**Tables S5 and S6**). Therefore, this data set provides an excellent opportunity to examine the generality of our theoretical predictions. After careful data selection (see **Materials and Methods** and **Supplementary Text**), we estimated $\alpha$, $\beta$, and $\gamma$ diversity (asymptotic species richness; **Materials and Methods**) at `r nrow(wsd_subset[[1]]) + nrow(wsd_subset[[2]])` watersheds (`r nrow(wsd_subset[[1]])` in Hokkaido and `r nrow(wsd_subset[[2]])` in the Midwest), each of which comprised $\ge$ 10 sites of presence-absence fish community data (a total of `r nrow(point_subset[[1]]) + nrow(point_subset[[2]])` sites). These watersheds are small enough (`r round(min(c(wsd_subset[[1]]$area, wsd_subset[[2]]$area)))` to `r round(max(c(wsd_subset[[1]]$area, wsd_subset[[2]]$area)))` km^2^) to assume that fishes can disperse therein at a multi-generation time scale [@comte2018], while posing challenges to traverse across watersheds (the ocean or lentic habitats; see **Materials and Methods** for watershed definition). We combined this data set with geospatial information, including watershed characteristics (watershed area and branching probability), climates (annual mean temperature and annual cumulative precipitation), and land use patterns (the fraction of agricultural land use and dam density). Using this data set, we assessed whether observed effects of the macro-scale factors are consistent across the two geographic regions by developing global and region-specific models. In the global model, we assumed that effects of ecosystem size (watershed area) and complexity (branching probability) are constant across the two regions (i.e., fixed slopes). Meanwhile, the region-specific model assumes region-specific slopes of ecosystem size and complexity (**Materials and Methods**). We compared the support of these competing models using the Bayes factor, a measure of the strength of evidence in favor of one model over the alternative. In our definition (**Materials and Methods**), the Bayes factor of $> 1$ indicates the support for the global model over the region-specific model.

Despite the substantial difference in fish fauna between the study regions, we found strong support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`): this result suggests the consistent effects of ecosystem size and complexity on $\gamma$ diversity across the two geographic regions. Furthermore, the observed patterns were closely aligned with our theoretical predictions (**Table 1 and Figure 4**). The estimated $\gamma$ diversity increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) across watersheds. These effects remained significant even after controlling for the potential influences of other environmental factors (precipitation, temperature, agricultural land use, and dam density) (**Table 1**). Our statistical analysis provides the first empirical evidence that ecosystem complexity regulates riverine biodiversity at the metacommunity level while conferring additional support for the ecosystem size effect on fish communities [@iwasaki2012].

Similarly, we found weak to moderate supports for the global models of $\alpha$ and $\beta$ diversity (Bayes factor = `r round(fit$alpha$bayes_factor, 1)` and `r round(fit$beta$bayes_factor, 1)`, respectively). In both regions, $\beta$ diversity responded significantly to ecosystem size and complexity while $\alpha$ diversity showing a weaker or a vague response to these ecosystem properties (**Table 1 and Figure 4**). In our simulations, this pattern has emerged under the scenarios with dispersal limitation, which elegantly matches the previous observations of stream fish movement. Field studies (e.g., mark-recapture) recurrently revealed the restricted movement of stream fish, typically limited to several tens to hundreds of meters at an annual scale [@radinger2014]. The reciprocal agreement of theoretical and empirical patterns provides indirect but convincing evidence that dispersal limitation plays a key role in driving the associations between $\gamma$ diversity and ecosystem properties in rivers.

@@

# Implications for biodiversity conservation

The emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation [@poff2007], habitat fragmentation [@grill2019], and stream burial [@elmore2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research [@koning2020]. While large reserves or restoration sites are undoubtedly important, our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin [@koning2020].

While the prevailing evidence supports the importance of ecosystem size in scaling species diversity [@hubbell2001; @triantis2012; @iwasaki2012], ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial [@turner2015] and aquatic ecosystems [@rodríguez-iturbe2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings of the dual control of biodiversity should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. The dual dimensional approach - size and complexity - broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.
